---
title: "CRISPR sorting screen analysis method comparison"
author: "Carl de Boer"
date: "5/10/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CRISPR sorting screen analysis method comparison}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CRISPR sort-screen evaluation test

This document includes both the code and results for the evaluation metrics used in the MAUDE paper. This should allow fairly easy extension to allow new approaches and datasets. 

```{r load libraries, results="hide", message = FALSE, warning=FALSE}
library("ggplot2")
library(reshape)
library(cowplot)
library(MAUDE)
library(openxlsx)
library(edgeR)
library(DESeq2)
```

I forgot to set a seed when doing this for the publication, and there is an element of stochasticity in the evaluation, and so the final results are likely to differ a little from the publication, especially for the datasets with less signal/noise; if you run the stochastic steps multiple times, set a different seed, or don't set the seed at all, they may also differ from this tutorial
```{r set seed}
set.seed(35263377)
```

### CD69 screen
Load the CD69 data and process it.
```{r Load and parse CD69 data}
#a mapping to unify bin names from Simeonov data
binmapBack = list("baseline" = "baseline", "low"="low", "medium"="medium","high"="high","back_" = "NS",
                  "baseline_" = "baseline", "low_"="low", "medium_"="medium", "high_"="high", 
                  "A"="baseline", "B"="low", "E" = "medium", "F"="high")

#this comes from manually reconstructing the CD69 density curve from extended data figure 1a (Simeonov et al)
binBoundsCD69 = data.frame(Bin = c("A","F","B","E"), 
                           fraction = c(0.65747100, 0.02792824, 0.25146688, 0.06313389), 
                           stringsAsFactors = FALSE) 
fractionalBinBounds = makeBinModel(binBoundsCD69[c("Bin","fraction")])
fractionalBinBounds = rbind(fractionalBinBounds, fractionalBinBounds)
fractionalBinBounds$screen = c(rep("1",6),rep("2",6));
#only keep bins A,B,E,F
fractionalBinBounds = fractionalBinBounds[fractionalBinBounds$Bin %in% c("A","B","E","F"),]
fractionalBinBounds$Bin = unlist(binmapBack[fractionalBinBounds$Bin]);

#load data
cd69OriginalResults = read.xlsx('https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5675716/bin/NIHMS913084-supplement-supplementary_table_1.xlsx')
cd69OriginalResults$NT = grepl("negative_control", cd69OriginalResults$gRNA_systematic_name)
cd69OriginalResults$pos = cd69OriginalResults$PAM_3primeEnd_coord;
cd69OriginalResults = unique(cd69OriginalResults)
cd69CountData = melt(cd69OriginalResults, id.vars = c("pos","NT","gRNA_systematic_name"))
cd69CountData = cd69CountData[grepl(".count$",cd69CountData$variable),]
cd69CountData$theirBin = gsub("CD69(.*)([12]).count","\\1",cd69CountData$variable)
cd69CountData$screen = gsub("CD69(.*)([12]).count","\\2",cd69CountData$variable)
cd69CountData$reads= as.numeric(cd69CountData$value); cd69CountData$value=NULL;
# convert their bin to one that is consistent
cd69CountData$Bin = unlist(binmapBack[cd69CountData$theirBin]); 
binReadMat = data.frame(cast(cd69CountData[!is.na(cd69CountData$pos) | cd69CountData$NT,], pos+gRNA_systematic_name+NT+screen ~ Bin, value="reads"))
```


I wanted to confirm that this is how they calculated logFC in Simeonov et al. The plot below should form a straight line on y=x.
```{r calc logFC}
#confirm how to calc log2FC:
cd69OriginalResults$l2fc.vsbg1 = log2((1+ cd69OriginalResults$CD69high_1.count.norm)/(1+cd69OriginalResults$CD69back_1.count.norm))
cd69OriginalResults$l2fc.vsbg2 = log2((1+ cd69OriginalResults$CD69high_2.count.norm)/(1+cd69OriginalResults$CD69back_2.count.norm))
cd69OriginalResults$l2fc.hilo1 = log2((1+ cd69OriginalResults$CD69high_1.count.norm)/(1+cd69OriginalResults$CD69baseline_1.count.norm))
cd69OriginalResults$l2fc.hilo2 = log2((1+ cd69OriginalResults$CD69high_2.count.norm)/(1+cd69OriginalResults$CD69baseline_2.count.norm))
#confirm that how we just calculated log2 fc is the same as originally calculated
p=ggplot(cd69OriginalResults, aes(x=high2.l2fc,y= l2fc.vsbg2)) + geom_point(); print(p)
```

Run the various analysis methods on the CD69 data.
```{r run methods, results="hide", message = FALSE, warning=FALSE}
#edgeR
x <- unique(cd69OriginalResults[c("gRNA_systematic_name","CD69baseline_1.count","CD69baseline_2.count", 
                                  "CD69high_1.count", "CD69high2.count")])
row.names(x) = x$gRNA_systematic_name; x$gRNA_systematic_name=NULL;
x=as.matrix(x)
group <- factor(c(1,1,2,2))
y <- DGEList(counts=x,group=group)
y <- calcNormFactors(y)
design <- model.matrix(~group)
y <- estimateDisp(y,design)
#To perform likelihood ratio tests:
fit <- glmFit(y,design)
lrt <- glmLRT(fit,coef=2)
edgeRGuideLevel = topTags(lrt, n=nrow(x))@.Data[[1]]
edgeRGuideLevel$gRNA_systematic_name = row.names(edgeRGuideLevel);
edgeRGuideLevel$metric = edgeRGuideLevel$logFC;
edgeRGuideLevel$significance = -log(edgeRGuideLevel$PValue);
edgeRGuideLevel$method="edgeR";

#DESeq2
deseqGroups = data.frame(bin=factor(c(1,1,2,2)));
row.names(deseqGroups) = c("CD69baseline_1.count","CD69baseline_2.count", 
                           "CD69high_1.count", "CD69high2.count");
dds <- DESeqDataSetFromMatrix(countData = x,colData = deseqGroups, design= ~ bin)
dds <- DESeq(dds)
res <- results(dds, name=resultsNames(dds)[2])

deseqGuideLevel = as.data.frame(res@listData)
deseqGuideLevel$gRNA_systematic_name =res@rownames;
deseqGuideLevel$metric = deseqGuideLevel$log2FoldChange;
deseqGuideLevel$significance = -log(deseqGuideLevel$pvalue);
deseqGuideLevel$method="DESeq2";

#MAUDE
guideLevelStatsCD69 = findGuideHitsAllScreens(unique(binReadMat["screen"]), binReadMat, fractionalBinBounds, sortBins = c("baseline","high","low","medium"), unsortedBin = "NS")
guideLevelStatsCD69$chr="chr12"
guideLevelStatsCastCD69 = cast(guideLevelStatsCD69, gRNA_systematic_name + pos+NT ~ screen, value="Z")
names(guideLevelStatsCastCD69)[ncol(guideLevelStatsCastCD69)-1:0]=c("s1","s2")
guideLevelStatsCastCD69$significance = apply(guideLevelStatsCastCD69[c("s1","s2")],1, combineZStouffer)
guideLevelStatsCastCD69$metric=apply(guideLevelStatsCastCD69[c("s1","s2")],1, mean)
guideLevelStatsCastCD69$method = "MAUDE"

#Two log fold change methods
cd69OriginalResultsHiLow = cd69OriginalResults[c("gRNA_systematic_name","l2fc.hilo1","l2fc.hilo2")]
cd69OriginalResultsVsBG = cd69OriginalResults[c("gRNA_systematic_name","l2fc.vsbg1","l2fc.vsbg2")]
cd69OriginalResultsHiLow$significance = apply(cd69OriginalResultsHiLow[2:3], 1, FUN = mean)
cd69OriginalResultsHiLow$metric = apply(cd69OriginalResultsHiLow[2:3], 1, FUN = mean)
cd69OriginalResultsVsBG$significance = apply(cd69OriginalResultsVsBG[2:3], 1, FUN = mean)
cd69OriginalResultsVsBG$metric = apply(cd69OriginalResultsVsBG[2:3], 1, FUN = mean)
cd69OriginalResultsHiLow$method="logHivsLow"
cd69OriginalResultsVsBG$method="logVsUnsorted"
```

Compile the results from the various methods.
```{r compile results}
# predictions
allResults = rbind(cd69OriginalResultsVsBG[c("method","gRNA_systematic_name","significance","metric")],
                   cd69OriginalResultsHiLow[c("method","gRNA_systematic_name","significance","metric")],
                   deseqGuideLevel[c("method","gRNA_systematic_name","significance","metric")], 
                   edgeRGuideLevel[c("method","gRNA_systematic_name","significance","metric")],
                   guideLevelStatsCastCD69[c("method","gRNA_systematic_name","significance","metric")]) 


allResults = merge(allResults, cd69OriginalResults[c("gRNA_systematic_name","NT","pos")],
                   by="gRNA_systematic_name")
allResults = allResults[!is.na(allResults$pos) | allResults$NT,]
allResults$promoter  = allResults$pos <= 9913996 & allResults$pos >= 9912997
allResults$gID = allResults$gRNA_systematic_name; allResults$gRNA_systematic_name=NULL;
allResults$locus ="CD69"
allResults$type ="CRISPRa"
allResults$celltype ="Jurkat"
```

### TNFAIP3
Load and parse the TNFAIP3 data from Ray et al.
```{r input and parse TNFAIP3 data}

##read in TNFAIP3 data
binFractionsA20 = read.table(textConnection(readLines(gzcon(url("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE136nnn/GSE136693/suppl/GSE136693_20190828_CRISPR_FF_bin_fractions.txt.gz")))), 
                             sep="\t", stringsAsFactors = FALSE, header = TRUE)
CRISPRaCountsA20 = read.table(textConnection(readLines(gzcon(url("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE136nnn/GSE136693/suppl/GSE136693_20190828_CRISPRa_FF_countMatrix.txt.gz")))), 
                              sep="\t", stringsAsFactors = FALSE, header = TRUE)
CRISPRiCountsA20 = read.table(textConnection(readLines(gzcon(url("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE136nnn/GSE136693/suppl/GSE136693_20190828_CRISPRi_FF_countMatrix.txt.gz")))), 
                              sep="\t", stringsAsFactors = FALSE, header = TRUE)
crispraGuides = read.table(textConnection(readLines(gzcon(url("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE136nnn/GSE136693/suppl/GSE136693_20180205_selected_CRISPRa_guides_seq.txt.gz")))), 
                           sep="\t", stringsAsFactors = FALSE, header = TRUE)
crispraGuides$seq = gsub("(.*)(.{3})","\\1",crispraGuides$seq.w.PAM)
crispraGuides$pos = round((as.numeric(gsub("([0-9]+):([0-9]+)-([0-9]+):([+-])","\\2",
                                           crispraGuides$guideID))+
                             as.numeric(gsub("([0-9]+):([0-9]+)-([0-9]+):([+-])","\\3",
                                             crispraGuides$guideID)))/2)

CRISPRaCountsA20 = merge(CRISPRaCountsA20, unique(crispraGuides[c("seq","pos")]), by=c("seq"), all.x=TRUE)

CRISPRiCountsA20$pos = round((as.numeric(gsub("([0-9]+):([0-9]+)-([0-9]+):([+-])","\\2",
                                              CRISPRiCountsA20$gID))+
                                as.numeric(gsub("([0-9]+):([0-9]+)-([0-9]+):([+-])","\\3",
                                              CRISPRiCountsA20$gID)))/2)

binFractionsA20$expt = paste(binFractionsA20$celltype, binFractionsA20$CRISPRType,sep="_")

#merge CRISPRi and CRISPRa
A20CountData = melt(CRISPRaCountsA20, id.vars=c("seq","elementClass","element","NT","gID","pos"))
A20CountData$CRISPRType="CRISPRa";
A20CountData2 = melt(CRISPRiCountsA20, id.vars=c("seq","elementClass","element","NT","gID","pos"))
A20CountData2$CRISPRType="CRISPRi";
A20CountData = rbind(A20CountData, A20CountData2)
rm('A20CountData2');
A20CountData$count = A20CountData$value; A20CountData$value=NULL;
A20CountData$sample = A20CountData$variable; A20CountData$variable=NULL;
A20CountData$bin = gsub("(.*)_(.*)_(.*)", "\\3", A20CountData$sample);
A20CountData$screen = gsub("(.*)_(.*)_(.*)", "\\2", A20CountData$sample);
A20CountData$celltype = gsub("(.*)_(.*)_(.*)", "\\1", A20CountData$sample);
A20CountData$expt = paste(A20CountData$celltype, A20CountData$CRISPRType,sep="_")
```

Run each method on each experiment from Ray et al. 
```{r start evaluation and run on TNFAIP3, results="hide", message = FALSE, warning=FALSE}
#combine CD69 metrics 
# Pearson's r between replicates ## metricsBoth will contain all our evaluation metrics
metricsBoth = 
  data.frame(method=c("MAUDE","logHivsLow","logVsUnsorted"), metric="r",which="replicate_correl",
             value=c(cor(guideLevelStatsCastCD69$s1[!guideLevelStatsCastCD69$NT],
                         guideLevelStatsCastCD69$s2[!guideLevelStatsCastCD69$NT]), 
                     cor(cd69OriginalResults$l2fc.hilo1[!cd69OriginalResults$NT],
                         cd69OriginalResults$l2fc.hilo2[!cd69OriginalResults$NT]),
                     cor(cd69OriginalResults$l2fc.vsbg1[!cd69OriginalResults$NT],
                         cd69OriginalResults$l2fc.vsbg2[!cd69OriginalResults$NT])),
             sig=c(cor.test(guideLevelStatsCastCD69$s1[!guideLevelStatsCastCD69$NT],
                            guideLevelStatsCastCD69$s2[!guideLevelStatsCastCD69$NT])$p.value, 
                   cor.test(cd69OriginalResults$l2fc.hilo1[!cd69OriginalResults$NT],
                            cd69OriginalResults$l2fc.hilo2[!cd69OriginalResults$NT])$p.value,
                   cor.test(cd69OriginalResults$l2fc.vsbg1[!cd69OriginalResults$NT],
                            cd69OriginalResults$l2fc.vsbg2[!cd69OriginalResults$NT])$p.value),
             locus="CD69",type="CRISPRa",celltype="Jurkat",stringsAsFactors = FALSE)

allResultsBoth = allResults; #combined results (significance, effect sizes) for CD69 and A20 screens
for (e in unique(binFractionsA20$expt)){
  curCelltype = gsub("(.*)_(.*)", "\\1", e);
  curtype = gsub("(.*)_(.*)", "\\2", e);
  curA20CountData = unique(A20CountData[A20CountData$expt==e,
                                        c("seq","pos", "NT","gID","count","screen","bin")])
  curA20CountDataTotals = cast(curA20CountData, screen +bin~ ., fun.aggregate = sum, value="count")
  names(curA20CountDataTotals)[3] = "total";
  curA20CountData = merge(curA20CountData, curA20CountDataTotals, by=c("screen","bin"))
  curA20CountData$CPM = curA20CountData$count/curA20CountData$total * 1E6;
  curCPMMat = cast(curA20CountData, seq + NT + gID + screen + pos ~ bin, value="CPM")
  curCPMMat$l2fc_hilo = log2((1+curCPMMat$F)/(1+curCPMMat$A))
  if(curtype=="CRISRPi"){
    curCPMMat$l2fc_vsbg = log2((1+curCPMMat$NS)/(1+curCPMMat$A))
  }else{ #CRISPRa
    curCPMMat$l2fc_vsbg = log2((1+curCPMMat$F)/(1+curCPMMat$NS))
  }
  curBins = as.data.frame(melt(binFractionsA20[binFractionsA20$expt==e,], 
                               id.vars = c("celltype","screen","CRISPRType","expt")))
  names(curBins)[ncol(curBins) - (1:0)] = c("Bin","fraction")
  curBins2 = data.frame();
  for (s in unique(curBins$screen)){
    curBins3 = makeBinModel(curBins[curBins$screen==s,c("Bin","fraction")])
    curBins3$screen = s;
    curBins2 = rbind(curBins2, curBins3)
  }
  curBins2$Bin = as.character(curBins2$Bin);
  curCountMat = cast(curA20CountData, seq + NT + gID +pos + screen ~ bin, value="count")
  guideLevelStats = findGuideHitsAllScreens(experiments = unique(curCountMat["screen"]), 
                                            countDataFrame = curCountMat, binStats = curBins2, 
                                            sortBins = c("A","B","C","D","E","F"), unsortedBin = "NS", 
                                            negativeControl="NT")
  
  guideLevelStatsCast = cast(guideLevelStats, gID + pos+NT ~ screen, value="Z")
  #names(guideLevelStatsCast)[4:ncol(guideLevelStatsCast)]=sprintf("s%i", 1:(ncol(guideLevelStatsCast)-3))
  
  maudeZs = guideLevelStatsCast;
  
  guideLevelStatsCast$significance = apply(maudeZs[unique(curA20CountData$screen)],1, combineZStouffer)
  guideLevelStatsCast$metric=apply(maudeZs[unique(curA20CountData$screen)],1, mean)
  guideLevelStatsCast$method = "MAUDE"
  
  ### EdgeR
  library(edgeR)
  x= cast(unique(curA20CountData[curA20CountData$bin %in% c("A","F"), c("bin","gID","screen","count")]), gID ~ screen + bin, value="count")
  row.names(x) = x$gID; x$gID=NULL;
  x=as.matrix(x)
  group = grepl("_F",colnames(x))+1
  group <- factor(group)
  y <- DGEList(counts=x,group=group)
  y <- calcNormFactors(y)
  design <- model.matrix(~group)
  y <- estimateDisp(y,design)
  #To perform likelihood ratio tests:
  fit <- glmFit(y,design)
  lrt <- glmLRT(fit,coef=2)
  edgeRGuideLevel = topTags(lrt, n=nrow(x))@.Data[[1]]
  
  edgeRGuideLevel$gID = row.names(edgeRGuideLevel);
  edgeRGuideLevel$metric = edgeRGuideLevel$logFC;
  edgeRGuideLevel$significance = -log(edgeRGuideLevel$PValue);
  edgeRGuideLevel$method="edgeR";
  
  ### DEseq
  library(DESeq2)
  deseqGroups = data.frame(bin=group);
  row.names(deseqGroups) = colnames(x);
  dds <- DESeqDataSetFromMatrix(countData = x,colData = deseqGroups, design= ~ bin)
  dds <- DESeq(dds)
  #resultsNames(dds) # lists the coefficients
  res <- results(dds, name=resultsNames(dds)[2])
  stopifnot(resultsNames(dds)[1]=="Intercept")
  deseqGuideLevel = as.data.frame(res@listData)
  deseqGuideLevel$gID =res@rownames;
  deseqGuideLevel$metric = deseqGuideLevel$log2FoldChange;
  deseqGuideLevel$significance = -log(deseqGuideLevel$pvalue);
  deseqGuideLevel$method="DESeq2";
  
  curLRHiLow = cast(unique(curCPMMat[c("gID","NT","screen","l2fc_hilo")]), gID + NT ~ screen, value="l2fc_hilo")
  curLRVsBG = cast(unique(curCPMMat[c("gID","NT","screen","l2fc_vsbg")]), gID + NT ~ screen, value="l2fc_vsbg")
  numSamples = ncol(curLRHiLow)-2;
  sampleNames = unique(curCPMMat$screen)
  
  curLRVsBG$significance = apply(curLRVsBG[3:(numSamples+2)], MARGIN = 1, FUN = mean);
  curLRVsBG$metric = apply(curLRVsBG[3:(numSamples+2)], MARGIN = 1, FUN = mean);
  curLRVsBG$method="logVsUnsorted"
  curLRHiLow$significance = apply(curLRHiLow[3:(numSamples+2)], MARGIN = 1, FUN = mean);
  curLRHiLow$metric = apply(curLRHiLow[3:(numSamples+2)], MARGIN = 1, FUN = mean);
  curLRHiLow$method="logHivsLow"
  
  #compile results for A20
  curResults = rbind(unique(curLRHiLow[c("method","gID","significance","metric")]),
                     unique(curLRVsBG[c("method","gID","significance","metric")]),
                     deseqGuideLevel[c("method","gID","significance","metric")], 
                     edgeRGuideLevel[c("method","gID","significance","metric")],
                     unique(guideLevelStatsCast[c("method","gID","significance","metric")]))
  
  curResults = merge(curResults, unique(curCPMMat[c("gID","NT","pos")]), by="gID")
  curResults = curResults[!is.na(curResults$pos) | curResults$NT,]
  curResults$promoter  = grepl("TNFAIP3", curResults$gID) | 
    (curResults$pos <= 138189439 & curResults$pos >= 138187040) # chr6:138188077-138188379;138187040
  
  #append the current results to all
  curResults$locus ="TNFAIP3"
  curResults$type =curtype
  curResults$celltype =curCelltype
  allResultsBoth = rbind(allResultsBoth, curResults);
  
  # (1) similarity between the effect sizes estimated per replicate, 
  corLRHiLow = cor(curLRHiLow[!curLRHiLow$NT, 3:(3+numSamples-1)])
  corLRVsBG = cor(curLRVsBG[!curLRVsBG$NT, 3:(3+numSamples-1)])
  maudeZCors = cor(maudeZs[!maudeZs$NT, 4:ncol(maudeZs)])
  
  maudeCorP=1
  maudeCorR=-1
  corLRHiLowP=1
  corLRHiLowR=-1
  corLRVsBGP=1;
  corLRVsBGR=-1
  #select the best inter-replicate correlation for each of the three approaches for which this is possible
  for(i in 1:(length(sampleNames)-1)){ 
    for(j in (i+1):length(sampleNames)){ 
      curR = cor(maudeZs[!maudeZs$NT, sampleNames[i]], maudeZs[!maudeZs$NT, sampleNames[j]])
      curP = cor.test(maudeZs[!maudeZs$NT, sampleNames[i]], maudeZs[!maudeZs$NT, sampleNames[j]])$p.value
      if (maudeCorR < curR){
        maudeCorR = curR;
        maudeCorP = curP;
      }
      curR = cor(curLRVsBG[!curLRVsBG$NT, sampleNames[i]], curLRVsBG[!curLRVsBG$NT, sampleNames[j]])
      curP = cor.test(curLRVsBG[!curLRVsBG$NT, sampleNames[i]], 
                      curLRVsBG[!curLRVsBG$NT, sampleNames[j]])$p.value
      if (corLRVsBGR < curR){
        corLRVsBGR = curR;
        corLRVsBGP = curP;
      }
      curR = cor(curLRHiLow[!curLRHiLow$NT, sampleNames[i]], curLRHiLow[!curLRHiLow$NT, sampleNames[j]])
      curP = cor.test(curLRHiLow[!curLRHiLow$NT, sampleNames[i]], 
                      curLRHiLow[!curLRHiLow$NT, sampleNames[j]])$p.value
      if (corLRHiLowR < curR){
        corLRHiLowR = curR;
        corLRHiLowP = curP;
      }
    }
  }
  metricsBoth = rbind(metricsBoth, 
                      data.frame(method=c("MAUDE","logHivsLow","logVsUnsorted"), 
                                 metric="r",which="replicate_correl",
                                 value=c(maudeCorR, corLRHiLowR, corLRVsBGR),
                                 sig=c(maudeCorP, corLRHiLowP, corLRVsBGP), 
                                 locus ="TNFAIP3", type =curtype, celltype =curCelltype, 
                                 stringsAsFactors = FALSE))
}
```

Here are some functions that I will make use of below.
```{r some useful functions}
#Some useful functions
ranksumROC = function(x,y,na.rm=TRUE,...){
  if (na.rm){
    x=na.rm(x);
    y=na.rm(y);
  }
  curTest = wilcox.test(x,y,...);
  curTest$AUROC = (curTest$statistic/length(x))/length(y)
  return(curTest)
}
na.rm = function(x){ x[!is.na(x)]}
```

### Evaluation
The inter-replicate correlations were calculated in the sections above and are stored in `metricsBoth`. Below, the two remaining metrics are calculated (distinguishing promoter vs other targeting guides, and adjacent vs randomly paired guides).
```{r evaluate methods part 2 and 3}
# (1) similarity between the effect sizes estimated per replicate, 
# (above)
metricsBoth$significant= metricsBoth$sig < 0.01;

# Other evaluation metrics
allExpts = unique(allResultsBoth[c("celltype","locus","type")])
for (ei in 1:nrow(allExpts)){
  curCelltype = allExpts$celltype[ei]
  curtype = allExpts$type[ei];
  curLocus = allExpts$locus[ei];
  
  curResults = allResultsBoth[allResultsBoth$celltype==curCelltype & 
                                allResultsBoth$type==curtype & allResultsBoth$locus==curLocus,]
  for(m in unique(curResults$method)){
    curData = curResults[curResults$method==m & !curResults$NT,]
    
    # (2) similarity in effect size between adjacent guides
    curData = curData[order(curData$pos),]
    guideEffectDistances = 
      data.frame(method = m, random=FALSE, 
                 difference = abs(curData$metric[2:nrow(curData)] - curData$metric[1:(nrow(curData)-1)]), 
                 dist =abs(curData$pos[2:nrow(curData)] - curData$pos[1:(nrow(curData)-1)]), 
                 stringsAsFactors = FALSE)
    guideEffectDistances = guideEffectDistances[guideEffectDistances$dist < 100,]
    guideEffectDistances$dist=NULL;
    ### changed 10 in next line to 100 to make this more robust
    curData = curData[sample(nrow(curData), size = nrow(curData)*100, replace = TRUE),]
    guideEffectDistances = 
      rbind(guideEffectDistances,
            data.frame(method = m, random=TRUE,
                       difference = abs(curData$metric[2:nrow(curData)] - 
                                          curData$metric[1:(nrow(curData)-1)]), stringsAsFactors = FALSE));
    # random should have more different than adjacent
    curRS = ranksumROC(guideEffectDistances$difference[guideEffectDistances$method==m &
                                                         guideEffectDistances$random],
                       guideEffectDistances$difference[guideEffectDistances$method==m &
                                                         !guideEffectDistances$random]) 
    metricsBoth = rbind(metricsBoth, data.frame(method=m, metric="AUROC-0.5",
                                                which="adjacent_vs_random",value = curRS$AUROC-0.5,
                                                locus=curLocus,celltype=curCelltype, type=curtype,
                                                sig=curRS$p.value, significant = curRS$p.value < 0.01))
    
    # (3) ability to distinguish promoter-targeting guides from other guides. 
    if (curtype=="CRISPRi" & !(m %in% c("edgeR","DESeq2"))){ 
      # edgeR and DESeq2 are reversed for CRISPRi
      # non promoter should have higher effect than promoter (more -ve)
      curRS = ranksumROC(curResults$significance[curResults$method==m & !curResults$NT &
                                                   !curResults$promoter],
                         curResults$significance[curResults$method==m & !curResults$NT &
                                                   curResults$promoter]) 
    }else{
      # promoter should have larger effect than non-promoter
      curRS = ranksumROC(curResults$significance[curResults$method==m & !curResults$NT &
                                                   curResults$promoter],
                         curResults$significance[curResults$method==m & !curResults$NT &
                                                   !curResults$promoter]) 
    }
    metricsBoth = rbind(metricsBoth, data.frame(method=m, metric="AUROC-0.5", which="promoter_vs_T",
                                                value = curRS$AUROC-0.5, locus=curLocus, 
                                                celltype=curCelltype, type=curtype, sig=curRS$p.value,
                                                significant = curRS$p.value < 0.01))
  }
}

##compile all metrics; label the best in each test and whether any tests were significant (P<0.01)
metricsBoth2 = metricsBoth;
metricsBoth2$method = factor(as.character(metricsBoth2$method), 
                             levels = c("logVsUnsorted","logHivsLow","DESeq2","edgeR","MAUDE"))
metricsBoth2Best = cast(metricsBoth2, which + locus + type+celltype ~ ., value="value", 
                        fun.aggregate = max)
names(metricsBoth2Best)[ncol(metricsBoth2Best)] = "best"
metricsBoth2 = merge(metricsBoth2, metricsBoth2Best, by = c("which","locus","type","celltype"))
metricsBoth2AnySig = cast(metricsBoth2[colnames(metricsBoth2)!="value"], 
                          which + locus + type+celltype ~ ., value="significant", fun.aggregate = any)
names(metricsBoth2AnySig)[ncol(metricsBoth2AnySig)] = "anySig"
metricsBoth2 = merge(metricsBoth2, metricsBoth2AnySig, by = c("which","locus","type","celltype"))
metricsBoth2$isBest = metricsBoth2$value==metricsBoth2$best;
metricsBoth2$isBestNA = metricsBoth2$isBest;
metricsBoth2$isBestNA[!metricsBoth2$isBestNA]=NA;
metricsBoth2$pctOfMax = metricsBoth2$value/metricsBoth2$best * 100;

#fill in NAs for edgeR and DESeq2 which cannot have inter-replicate correlations
temp = metricsBoth2[metricsBoth2$metric=="r",]
temp = temp[1:2,];
temp$method = c("edgeR","DESeq2")
temp$value=NA; temp$isBest=NA; temp$significant=FALSE; temp$pctOfMax=NA; temp$isBestNA=NA;
metricsBoth2 = rbind(metricsBoth2, temp)
```

Finally make the graph with all evaluation metrics as shown in the MAUDE paper. 
```{r make graph, fig.width=10, fig.height=6}
#make the final evaluation graph
p1 = ggplot(metricsBoth2[metricsBoth2$which=="adjacent_vs_random",], 
            aes(x=method, fill=value, y=paste(locus,type,celltype))) + geom_tile() +
  geom_text(data=metricsBoth2[metricsBoth2$which=="adjacent_vs_random" & metricsBoth2$anySig,],
            aes(label="*",colour=isBestNA),show.legend = FALSE) +theme_bw() +
  scale_fill_gradient2(high="red", low="blue", mid="black") + 
  theme(legend.position="top", axis.text.x = element_text(hjust=1, angle=45), 
        axis.title.y = element_blank())+scale_colour_manual(values = c("green"), na.value=NA) + 
  scale_y_discrete(expand=c(0,0))+scale_x_discrete(expand=c(0,0))+ggtitle("Adjacent vs\nrandom guides");
p2 = ggplot(metricsBoth2[metricsBoth2$which=="promoter_vs_T",], 
            aes(x=method, fill=value, y=paste(locus,type,celltype))) + geom_tile() +
  geom_text(data=metricsBoth2[metricsBoth2$which=="promoter_vs_T" & metricsBoth2$anySig,],
            aes(label="*",colour=isBestNA),show.legend = FALSE) +theme_bw() +
  scale_fill_gradient2(high="red", low="blue", mid="black") + 
  theme(legend.position="top", axis.text.x = element_text(hjust=1, angle=45), 
        axis.title.y = element_blank())+scale_colour_manual(values = c("green"), na.value=NA)+
  scale_y_discrete(expand=c(0,0))+scale_x_discrete(expand=c(0,0))+
  ggtitle("Promoter vs other\ntargeting guides");
p3 = ggplot(metricsBoth2[metricsBoth2$which=="replicate_correl",], 
            aes(x=method, fill=value, y=paste(locus,type,celltype))) + geom_tile() +
  geom_text(data=metricsBoth2[metricsBoth2$which=="replicate_correl" & metricsBoth2$anySig,],
            aes(label="*",colour=isBestNA),show.legend = FALSE) +theme_bw() +
  scale_fill_gradient2(high="red", low="blue", mid="black") + 
  theme(legend.position="top", axis.text.x = element_text(hjust=1, angle=45), 
        axis.title.y = element_blank())+scale_colour_manual(values = c("green"), na.value=NA)+
  scale_y_discrete(expand=c(0,0))+scale_x_discrete(expand=c(0,0))+ggtitle("Replicate\ncorrelations");
g= plot_grid(p1,p2,p3, align = 'h', nrow = 1); print(g)
```

### Session info
```{r session info}
sessionInfo()
```

